---
title: "Surplus Production Model"
author: "Matt Robertson"
date: "`r format(Sys.time(), '%B %Y')`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Setup

I am going to make a fictitious fish population and track its dynamics in response to an increasing and then stabilizing amount of fishing pressure.

To do this, we first need to define the time over which we want to track this population’s dynamics as well as some basic components of this population’s biology.

I am going to run a simulation for 50 years, lets say from 2000 to 2050.

```{r time}
yrs<-seq(from=2000, to=2050)
nyears<-length(yrs)
```

When we imagine that a population grows following the dynamics modeled with a surplus production model, we only need to understand the population's maximum growth rate, $r_{max}$, and its carrying capacity, $K$.

We are going to start simply and expect that $r_{max} = 1$ and $K = 1000$. Additionally, we are going to assume that our population is at carrying capacity when our simulated fishery starts.

```{r growth}
K<-1000

rmax<-1
```

# The fishery

For the purposes of this simulation, I am going to assume that the fishery has followed a specific pattern of exploitation over time. In this case, catches will start at 0 in 2000 and steadily increase to a maximum value of 250 at which point it will level off and the fishery will maintain at a steady amount of fishing pressure until the end of the simulation.

```{r fishing}
effort<-rep(NA, nyears)
C<-rep(NA, nyears)
incF<-round(nyears*0.25,0)
effort[1:incF]<-seq(from=0, to=250, length=incF)
effort[(incF):nyears]<-250
for(i in 1:nyears){
C[i]<-effort[i] + rnorm(1, 0, 10)
}

plot(C~ yrs,pch=19, xlab="Year", ylab="Catch", las=1)
lines(C~ yrs)
```

# Simulation

Now that the species biology and fishery have been defined, we can run a theoretical simulation of how this fishery would impact this type of population. I have include some random variability into the annual fishing mortality and into the stock-recruitment relationship so no two simulations should ever be the same. I have hidden the code, because the details are not important for this class but we will look at how this population changes over time.

After the simulation runs, we can examine how the population actually changed over time. To start, we will look at total population biomass.

```{r sim, echo=FALSE}

B<-matrix(NA, nrow=nyears, ncol=1)
pop_growth<-matrix(NA, nrow=nyears, ncol=1)
B[1]<-K

for(i in 2:50){
  pop_growth[i-1]<-(rmax*B[i-1]*(1-(B[i-1]/K)))+rnorm(1,0,15)
  B[i]<-B[i-1]+pop_growth[i-1]-C[i-1]
}
```

We can see that the fishery seems to reduce our population size steadily at first but then the population stops declining and somewhat stabilizes.

```{r}
plot(B~yrs, pch=19, xlab="Year", ylab="Population Biomass", las=1)
lines(B~yrs)
```

We can then see how the population growth rate changed as biomass changed. When we do this, we see that the growth rate changes are almost inverse of the population size, where we see a steady increase followed by a plateau.

```{r}
plot(pop_growth~yrs, pch=19, xlab="Year", ylab="Annual Population Growth Rate", las=1)
lines(pop_growth~yrs)
```

If we compare how population growth has changed in relation to population size, we will see a very familiar trend appear. The red line indicates $B_{MSY}$, $K/2$, and the curve represents our expectations for how population growth will change with biomass.

```{r msy, echo=FALSE}
MSY<-K/2

simB<-seq(from=0, to=1000)
simgrowth<-(rmax*simB*(1-(simB/K)))

plot(simB, simgrowth, type="l", ylim=c(0,350), xlab="Population Biomass", ylab="Population Growth Rate", las=1)
points(B, pop_growth, pch=19)
abline(v=MSY, col="red")
```

We can also calculate what the catch at $MSY$ would be by using $MSY = \frac{rK}{4}$. In this case, that is 250.

```{r hmsy, echo=FALSE}
simC<-rmax*simB*(K-simB)/K

plot(C~B, xlim=c(0,1000), ylim=c(0,350), xlab="Population Biomass", ylab="Catch", las=1, pch=19)
lines(simC~simB)
abline(h=(rmax*K)/4, col="red")
```

